### 2017-11-16 manuscript.rmd
incorporated ssiebert comments, pdf not rebuilt

KMRR [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog can and has been applied to diverse attributes, including molecular sequence, biochemical function, and, as in the study considered here, expression. The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail. It has been applied to many traits of genes, from gene sequence to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

@@

We reanalyzed the KMRR study using phylogenetic comparative methods. We focused on one of the datasets included in their analyses, that of Brawand *et al.* 2011 [@Brawand:2011du]. This dataset is the best sampled in their analyses. It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and further considered here.

@@

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees %>% format(big.mark=",")` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned) %>% format(big.mark=",")` gene trees passed taxon sampling criteria (`r min_genes_with_expression %>% format(big.mark=",")` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)  %>% format(big.mark=",")` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow() %>% format(big.mark=",")` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow() %>% format(big.mark=",")` speciation nodes (Tables S1, S2). One of these trees is presented in *SI Appendix* Fig. 1 to demonstrate the analysis.

@@

We did not find increased evolutionary change in expression following duplication events as compared to speciation events (Fig. 2*B*). The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation ($P=`r wilcox_test_result`$). Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system. This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; Posada:2004ix] (*SI Appendix*, *SI Appendix* Fig. 2).

We next examined the possibility that ascertainment biases were differentially impacting the inference of expression evolution following duplication and speciation events. Such a bias might obscure support for the ortholog conjecture. We focused on two possible sources of bias, node depth and branch length. We found no evidence that either affected our results (*SI Appendix*, Fig. 3). We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees. This is important because it is expected that genes from separate species have a common ancestor older than the time at which the species diverged from each other [@Takahata:1989uh; @Degnan:2009hr]. There is also uncertainty associated with the timing of these speciation events. We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (*SI Appendix*).

An additional concern is that within-species expression variation is not considered when making comparisons of Tau across homologous genes. Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; @Felsenstein:2008cl; @Rohlfs:2015bd]. Since branch length describes expected trait variation [@Felsenstein:1985ua], within species variation can be modeled by extending the terminal branches of each gene phylogeny [@Ives:2007cw]. We modeled within-species variation by extending each terminal branch by 7 million years. This value was selected because it corresponds to the age of the clade Hominini (defined by the most recent common ancestor of humans and chimpanzees), the shallowest node in the phylogeny of examined species. This provides an opportunity to assess the impact that extreme within-species variation (*i.e.*, as large as the variance between humans and chimpanzees and their most recent common ancestor) would have on our findings. These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Fig. 4).

@@

To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Fig. 5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig. 5*B*), the paralog correlation coefficient increases from `r signif( paralog_r$estimate, 2)` to `r signif( paralog_subset_r$estimate, 2)` ($P`r format_p(paralog_subset_r)`$). This is much closer to the ortholog correlation of `r signif( ortholog_r$estimate, 2) `.

@@

For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for Tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv]. We then collected the contrasts from all trees into a single table, along with other annotations including whether the node is a speciation or duplication event. This table, `nodes_contrast`, was then analyzed as described in the main text for the presented plots and tests.

@@

The gene trees considered here vary widely in their phylogenetic signal for Tau according to Blomberg's $K$ [@Blomberg:2003jg] (*SI Appendix*, Fig. S2). $K$ describes the tendancy for closely related tips to resemble each other, where $K=0$ indicates that close relatives resemble each other no more than they do distant relatives and $K=1$ indicates that the data have strong phylogenetic signal. To examine the possibility that gene trees with little phylogenetic signal were diluting support for the Ortholog Conjecture, we removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a $K$ cutoff for retained genes of greater than `r signif( k_thresh, 3)`. Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for Tau still do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_test_result_k %>% round( 3 )`$). This indicates that gene trees with little phylogenetic signal are not obscuring support for the Ortholog Conjecture in genes with greater phylogenetic signal.

Phylogenetic independent contrasts (PICs) are based on a Brownian Motion (BM) model of character evolution [@Felsenstein:1985ua]. PICs could fail to reject the null hypothesis if BM is not a good model for the evolution of Tau on the gene phylogenies. We therefore compared the fit of BM to the more complex Ornstein Uhlenbeck (OU) model of evolution with the Akaike Information Criterion (AIC) [@Akaike:1974ih; Posada:2004ix].

@@

The independent contrast across a node is the amount of change observed between the daughter nodes, scaled by the expected variance [@Felsenstein:1985ua]. The expected variance is principally determined by the lengths of the branches leading from the node to these daughters. The shorter the total length of the two branches leading to daughter nodes, the larger the contrast for a given observed difference. This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Fig. 3*C*). This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated. We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysis does not reject the null hypothesis either (*SI Appendix*, Fig. 3*D*), indicating that branch length bias is not responsible for the result.

>>>>

KMRR [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture can and has been applied to diverse attributes, including molecular sequence, biochemical function, and, as in the study considered here, expression. It has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail. It has been applied to many traits of genes, from gene sequence to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

@@

We reanalyzed the KMRR study using phylogenetic comparative methods. We focused on one of the datasets included in their analyses, that of Brawand *et al.* 2011 [@Brawand:2011du]. This dataset is the best sampled in their analyses. It has gene expression data for six organs across ten species (nine mammals and one bird), eight of which were analyzed by KMRR and are further considered here.

@@

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees %>% format(big.mark=",")` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned) %>% format(big.mark=",")` gene trees passed taxon sampling criteria (`r min_genes_with_expression %>% format(big.mark=",")` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)  %>% format(big.mark=",")` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow() %>% format(big.mark=",")` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow() %>% format(big.mark=",")` speciation nodes (Tables S1, S2). One of these trees is presented in *SI Appendix* Fig. S1 to demonstrate the analysis.

@@

We did not find increased evolutionary change in expression following duplication events as compared to speciation events (Fig. 2*B*). The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation ($P=`r wilcox_test_result`$). Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore finds no support for the ortholog conjecture in this system. This result holds when we exclude the genes with the lowest phylogenetic signal according to Blomberg's $K$ [@Blomberg:2003jg] and is robust to model selection [@Akaike:1974ih; @Posada:2004ix] (*SI Appendix*, *SI Appendix* Fig. S2).

We next examined the possibility that ascertainment biases were differentially impacting the inference of expression evolution following duplication and speciation events. Such a bias might obscure support for the ortholog conjecture. We focused on two possible sources of bias, node depth and branch length. We found no evidence that either affected our results (*SI Appendix*, Fig. S3). We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees. This is important because it is expected that genes from separate species have a common ancestor older than the time at which the species diverged from each other [@Takahata:1989uh; @Degnan:2009hr]. There is also uncertainty associated with the timing of these speciation events. We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (*SI Appendix*).

An additional concern is that within-species expression variation is not considered when making comparisons of Tau across homologous genes. Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; @Felsenstein:2008cl; @Rohlfs:2015bd]. Since branch length describes expected trait variation [@Felsenstein:1985ua], within species variation can be modeled by extending the terminal branches of each gene phylogeny [@Ives:2007cw]. We modeled within-species variation by extending each terminal branch by 7 million years. This value was selected because it corresponds to the age of the clade Hominini (defined by the most recent common ancestor of humans and chimpanzees), the shallowest node in the phylogeny of examined species. This provides an opportunity to assess the impact that extreme within-species variation (*i.e.*, as large as the variance between humans and chimpanzees and their most recent common ancestor) would have on our findings. These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Fig. S4).

@@

To test the hypothesis that ancient duplications that precede the oldest speciation event (*SI Appendix*, Fig. S5*A*) impact the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the duplication events the same age or younger than the oldest speciation event (*SI Appendix*, Fig. 5*B*), the paralog correlation coefficient increases from `r signif( paralog_r$estimate, 2)` to `r signif( paralog_subset_r$estimate, 2)` ($P`r format_p(paralog_subset_r)`$). This is much closer to the ortholog correlation of `r signif( ortholog_r$estimate, 2) `.

@@

For each node in the remaining calibrated trees, we calculated the phylogenetic independent contrast for Tau across its daughter branches with the `pic()` function in ape [@Paradis:2004dv]. Summary statistics on taxon and node sampling are presented in *SI Appendix* Tables S1-S2. We then collected the contrasts from all trees into a single data frame, along with other annotations including whether the node is a speciation or duplication event. This data frame, `nodes_contrast`, was then analyzed as described in the main text for the presented plots and tests.

@@

The gene trees considered here vary widely in their phylogenetic signal for Tau according to Blomberg's $K$ [@Blomberg:2003jg] (*SI Appendix*, Fig. S2). $K$ describes the tendency for closely related tips to resemble each other, where $K=0$ indicates that close relatives resemble each other no more than they do distant relatives and $K=1$ indicates that the data have strong phylogenetic signal. To examine the possibility that gene trees with little phylogenetic signal were diluting support for the Ortholog Conjecture, we removed the `r k_percentile` of genes with the lowest phylogenetic signal, corresponding to a $K$ cutoff for retained genes of greater than `r signif( k_thresh, 3)`. Repeating the phylogenetic analyses on this subset of genes with strongest phylogenetic signal for Tau still do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_test_result_k %>% round( 3 )`$). This indicates that gene trees with little phylogenetic signal are not obscuring support for the Ortholog Conjecture in genes with greater phylogenetic signal.

Phylogenetic independent contrasts (PICs) are based on a Brownian Motion (BM) model of character evolution [@Felsenstein:1985ua]. PICs could fail to reject the null hypothesis if BM is not a good model for the evolution of Tau on the gene phylogenies. We therefore compared the fit of BM to the more complex Ornstein Uhlenbeck (OU) model of evolution with the Akaike Information Criterion (AIC) [@Akaike:1974ih; @Posada:2004ix].

@@

The independent contrast across a node is the amount of change observed between the daughter nodes, scaled by the expected variance [@Felsenstein:1985ua]. The expected variance is principally determined by the lengths of the branches leading from the node to these daughters. The shorter the total length of the two branches leading to daughter nodes, the larger the contrast for a given observed difference. This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (*SI Appendix*, Fig. S3*C*). This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated. We therefore examined only the contrasts that fell within the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysis does not reject the null hypothesis either (*SI Appendix*, Fig. S3*D*), indicating that branch length bias is not responsible for the result.
